---
title: "Perform ModelArray.wrap whole-brain analysis and produce outputs files in the NBDC sandbox"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developing a function for ModelArray.wrap in the NBDC sandbox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## Introduction

The NBDC sandbox provides access to ABCD and HBCD hdf5 outputs for QSIrecon. This includes a lot of different measures such as mean diffusivity (md), fractional anisotropy (FA), and return to origin probability (RTOP). See diffusion MRI Data Processing details on the HBCD Data Release Docs site [here](https://docs.hbcdstudy.org/latest/instruments/mri/dmri/#data-processing) for an overview of metrics output by QSIRecon.

This walkthrough shows you how to use `ModelArray.wrap` to perform statistical analysis on hdf5 data.

We'll develop a function that computes the mean RTOP within each site, and then use `ModelArray.wrap` to run it across all voxels.

Note that for this tutorial, we will be using a phenotype file used specifically for [Cieslak et al. 2025](https://doi.org/10.1101/2025.11.10.687672), made available via a secure NIST SP 1800 system. This tutorial will be updated in the future to use tabulated HBCD Study data available via the NBDC Data Access Platform for consistency. 

### Step 1 Load the mapmri_rtop data in ModelArray

The first step is to load the HBCD RTOP HDF5 file and the accompanying CSV file.

After you've installed `ModelArray`, here's how we load the library and 
point it to the HDF5 file. 

You can choose the imaging data for your question by changing the `h5_path` to your preferred h5 file. You can also change the imaging measure loaded from the h5 file by changing `scalar_types`.

```{r load ModelArray}
library(ModelArray)
library(tidyr)
library(dplyr)
# Use the real path
h5_path <- "/shared/hackathon/working-area/hbcd-workshop/mapmri-rtop/mapmri_rtop.h5"
# create a ModelArray-class object:
rhdf5::h5ls(h5_path)
modelarray <- ModelArray(h5_path, scalar_types = c("mapmri_rtop"))
# let's check what's in it:
modelarray
```
You'll see:
```{.console}
ModelArray located at /shared/hackathon/working-area/hbcd-workshop/mapmri-rtop/mapmri_rtop.h5

  Source files:     XXXXX
  Scalars:          mapmri_rtop
  Analyses:
```


This shows that there are XXXXX source mapmri_rtop files in this `modelarray` you created.

You may take a look at what the scalar matrix looks like by using `scalars()`:
```{r access slots, eval = FALSE}
# scalar mapmri_rtop data:
scalars(modelarray)[["mapmri_rtop"]]
```
You'll see:
```{.console}
<YYYYY x XXXXX> DelayedMatrix object of type "double":
          file1.nii.gz ... fileXXXX.nii.gz
     [1,]                                                    2.323956   .                                                    3.401824
     [2,]                                                    2.927209   .                                                    2.428174
     [3,]                                                    3.211094   .                                                    2.479505
     [4,]                                                    2.463409   .                                                    3.760944
     [5,]                                                    2.460162   .                                                    2.912316
      ...                                                           .   .                                                           .
[327680,]                                                    2.832996   .                                                    3.066516
[327681,]                                                    2.669299   .                                                    2.912496
[327682,]                                                    2.555752   .                                                    2.510855
[327683,]                                                    2.336970   .                                                    2.472471
[327684,]                                                    2.256365   .                                                    2.293425
```
Rows are voxels `(n = YYYYY)`, and column names are the source filenames `(n = XXXXX)`.
Each value is a specific voxel's Mean Diffusivity from a nii.gz file.

### Step 2 Load the phenotypes CSV file

We then load the accompanying CSV file: `mapmri_rtop.csv`. To choose different behavioral or developmental measures for answering your questions, you will need to select a different csv file. To do this, change the `csv_path` under the code chunk titled `load csv file`. 

```{r load csv file, eval=FALSE}
csv_path <- "/shared/hackathon/working-area/hbcd-workshop/mapmri-rtop/mapmri_rtop.csv"
# load the CSV file:
phenotypes <- read.csv(csv_path)
nrow(phenotypes)
names(phenotypes)
```

This CSV file was used to build the HDF5 file: the included subjects/sessions match the h5 file exactly.
You will probably want to add additional phenotypes to this file, since it only contains dMRI QC data.

### Step 3 explore the data

With both `modelarray` and data frame `phenotypes` set up,
we can explore the data and prepare to write a function that we can use within `ModelArray.wrap` to compute what we want.
It can be tricky to figure out exactly what you want to compute and how to make sure your function returns the correct output.
We can use a convenience function from `ModelArray` to see what a dataframe looks like for a single voxel.

```{r check_one_element_df, eval=FALSE}
example_df <- exampleElementData(modelarray, scalar = "mapmri_rtop", i_element = 72110, phenotypes = phenotypes)
names(example_df)
```

This is a rich dataframe!
It contains the Mean Diffusivity for all subjects in the study at voxel 72110.
It also contains the subject ID, session ID, sex, age, and other covariates.

We can explore this single voxel like we would any dataframe: with data visualizations. Bar plot visualizations are effective ways of visualizing a single comparison. The code below can be customized to plot your preferred data. Change `x` from `session_id` to plot a different variable on the X axis. Change `y` to plot a different variable on the Y axis. Change `color` from `session_id` to color-code the bars by a different grouping variable.

```{r explore_one_element_df, eval=FALSE}
library(ggplot2)
ggplot(example_df, aes(x = session_id, y = mapmri_rtop, color = session_id)) +
  geom_boxplot() +
  theme_minimal()
```

Wow! That effect is super big!
Let's write a function that computes a mean Mean Diffusivity within each session.
We'll also compute an anova on session and get the statistics/p-values.
Note that this is a very simplified example and not nearly nuanced enough for a real analysis.

```{r anova_into_df_example, eval=FALSE}
session_anova <- example_df %>%
  {
    lm(mapmri_rtop ~ session_id, data = .)
  } %>%
  anova() %>%
  broom::tidy() %>%
  dplyr::mutate(
    partial_eta_sq = dplyr::if_else(
      term != "Residuals",
      sumsq / (sumsq + sumsq[term == "Residuals"]),
      NA_real_
    )
  ) %>%
  dplyr::filter(term != "Residuals") %>%
  dplyr::select(term, df, sumsq, meansq, statistic, p.value, partial_eta_sq) %>%
  tidyr::pivot_wider(
    names_from = term,
    values_from = c(df, sumsq, meansq, statistic, p.value, partial_eta_sq),
    names_glue = "{term}.{.value}"
  ) %>%
  tibble::as_tibble()

session_anova
```

In order to run this type of analysis on _all_ voxels, we need to write a function that can be used within `ModelArray.wrap`.
Absolutely critical here is that this function needs a single argument named `data`:

```{r full_function_example, eval=FALSE}
my_mean_mapmri_rtop_fun <- function(data) {
  session_means <- data %>%
    group_by(session_id) %>%
    summarize(mean_mapmri_rtop = mean(mapmri_rtop, na.rm = TRUE)) %>%
    pivot_wider(names_from = session_id, values_from = mean_mapmri_rtop, names_glue = "mapmri_rtop_{session_id}")
  session_anova <- data %>%
    {
      lm(mapmri_rtop ~ session_id, data = .)
    } %>%
    anova() %>%
    broom::tidy() %>%
    dplyr::mutate(
      partial_eta_sq = dplyr::if_else(
        term != "Residuals",
        sumsq / (sumsq + sumsq[term == "Residuals"]),
        NA_real_
      )
    ) %>%
    dplyr::filter(term != "Residuals") %>%
    dplyr::select(term, df, sumsq, meansq, statistic, p.value, partial_eta_sq) %>%
    tidyr::pivot_wider(
      names_from = term,
      values_from = c(df, sumsq, meansq, statistic, p.value, partial_eta_sq),
      names_glue = "{term}.{.value}"
    ) %>%
    tibble::as_tibble()
  bind_cols(session_means, session_anova)
}

my_mean_mapmri_rtop_fun(example_df)
```

This looks good.
We can test that the function will work in the `ModelArray` looping machinery by running it on a single voxel. Modelarray uses a `wrap` function to perform data analyses. A code chunk for running a single Modelarray comparison is reprinted below. Change `72110` to select a different imaging datapoint to analyze. Change `mampri_rtop` to the imaging datatype one wants to analyze. Change `my_mean_mapmri_rtop_fun` to select a different model for analyzing the data. Change `phenotypes` for selecting the non-imaging data to analyze. 

```{r test_one_element, eval=FALSE}
analyseOneElement.wrap(
  72110,
  my_mean_mapmri_rtop_fun,
  modelarray,
  phenotypes,
  "mapmri_rtop",
  num.subj.lthr = 2,
  flag_initiate = TRUE,
  on_error = "debug"
)
```
This will print the column names that will be retrieved from the function when it's run in the `ModelArray` looping machinery. This can be run on your own laptop! 

This piece can run it on all the voxels.

```{r run_wrapped_function, eval=FALSE}
res_wrap <- ModelArray.wrap(
  FUN = my_mean_mapmri_rtop_fun,
  data = modelarray,
  phenotypes = phenotypes,
  scalar = "mapmri_rtop",
  n_cores = 8
)
```

This will take a while.
Once it finiehes you can take a look at the results:

```{r head_results, eval=FALSE}
head(res_wrap)
```

```{.console}
# This needs to be replaced with actual output
  element_id mapmri_rtop_session_1 mapmri_rtop_session_2 mapmri_rtop_session_3 mapmri_rtop_session_4 session_id.df session_id.sumsq session_id.meansq session_id.statistic session_id.p.value session_id.partial_eta_sq
1          0              3.108483              3.009349            3.063119            2.692335             3         45.52142         15.173806             59.35110       5.688104e-37                0.07093606
2          1              2.618739              2.673533            2.494077            2.324791             3         24.51961          8.173205             28.88297       2.474739e-18                0.03582533
3          2              3.111822              3.056225            2.983709            2.812225             3         23.59899          7.866330             19.95448       9.059350e-13                0.02502795
4          3              3.669784              3.621715            3.054638            3.329743             3        186.55051         62.183503             83.32945       3.261715e-51                0.09682010
5          4              2.790135              2.729325            2.494522            2.444265             3         53.34644         17.782147             55.42033       1.336311e-34                0.06655069
6          5              2.996483              3.043458            2.785565            2.650955             3         40.25648         13.418828             41.60603       3.458560e-26                0.05080478
```


We now save the results data frame into the original h5 file:

```{r write results, eval=FALSE}
# CRITICAL NOTE: when we run this locally, we copy the h5 to our home directory and write the results into the copy.
# IDK if this actually works, creating a new h5 file with just the results. Will have to try it out.
writeResults("/path/to/a/new/h5/file.h5", df.output = res_wrap, analysis_name = "mapmri_rtop_anova_and_means")
```

Notice that the the analysis name specified in argument `analysis_name` will be used in `ModelArrayIO` in the next step when converting results back to nii.gz file format. 
It'll also be used as the prefix of the nii.gz files to be saved.


## Step 4 Convert results to nii.gz file format


### Step 4.1. Convert the statistical results into mif file format using ConFixel
We now use `ConFixel` to convert the results into a list of mif files:

```{.console}
$ voxelstats_write \
  --cohort-file /nbdc-sandbox/derivatives/qsirecon-modelarray/ABCC_mapmri_rtop_phenotypes.csv \
  --analysis-name mapmri_rtop_anova_and_means \
  --input-hdf5 /nbdc-sandbox/derivatives/qsirecon-modelarray/qsirecon-mapmri_rtop.h5 \
  --output-dir /where/you/want/to/put/your/niftis \
  --example-nifti /nbdc-sandbox/derivatives/qsirecon-modelarray/qsirecon-mapmri_rtop.nii.gz # Need to find the real path
```

After it's done, in the main folder `/where/you/want/to/put/your/niftis`, 
there will be a new folder called `mapmri_rtop_anova_and_means` and converted result `nii.gz` files are in this new folder: